Symmetric Circuits for Rank Gates

### 2018-03-10 sections/background.tex
I have been trying to improve the presentation of the translation of circuits into formulas. To this end, I have changed the notation we use to talk about the usual logical operators (e.g. AND, OR, NOT, etc). I have also added tried to simplify the writing, and better explain the approach taken when translating circuits into formulas. I am in the process of writing the conclusion, but there is so much to think about.

\subsection{Boolean Functions, Bases, and Circuits}
A (finite) \emph{Boolean function} is a function of the form $f: \{0,1\}^X
\rightarrow \{0,1\}$, for some finite index set $X$. We call a Boolean function
$f: \{0,1\}^X \rightarrow \{0,1\}$ \emph{symmetric} if for all $\sigma \in
\sym_X$ and $\vec{a} \in \{0,1\}^X$, $f (\vec{a} \sigma) = f(\vec{a})$. It
follows that a Boolean function is symmetric if, and only if, the output of the
function is entirely determined by the number of ones in the input. We say a
Boolean function $f : \{0,1\}^{*} \rightarrow \{0,1\}$ is \emph{symmetric} if
its restriction to inputs of size $n$ is symmetric for all $n$.

We depart slight from the conventional notion of a basis in this paper. We
define a \emph{Boolean basis} (or \emph{basis}) to be a set of finite Boolean
functions. We always denote a basis by $\BB$. It is common in the literature to
define a basis as a finite sequence of functions $f_1, \ldots , f_k:\{0,1\}^*
\rightarrow \{0,1\}$. In this paper we instead think of this sequence as
\emph{generating} a basis, and write $\BB = \langle f_1, \ldots, f_k \rangle$ to
denote the basis generated by these functions, i.e.\ $\BB$ consists of all the
Boolean functions $f_i[n] : \{0,1\}^n \rightarrow \{0,1\}$ where $n \in \nats$,
$i \in [k]$, and $f_i[n]$ is the restriction of $f_i$ to strings of length $n$.
In other words $\langle f_1, \ldots , f_k \rangle$ is the union of $\{f_i[n] : n
\in \nats\}$ for each $f_i$ in the generating sequence. We have $\AND, \OR,
\NAND, \MAJ : \{0,1\}^* \rightarrow \{0,1\}$, where $\AND, \OR, \NAND$ are the
usual (unbounded fan-in) logical operators, and $\MAJ$ is the majority function
(i.e.\ the function that outputs one if more than half of bits in the inputs are
one). We write $\BS := \langle \AND, \OR, \NAND\} \rangle$ to denote the
\emph{standard basis} and $\BM := \langle \AND, \OR, \NAND , \MAJ \rangle$ to
denote the \emph{majority basis}.

@@

\begin{definition}
  Let $(C_n)_{n \in \mathbb{N}}$ be a family of Boolean circuits. We say that
  $(C_n)_{n \in \mathbb{N}}$ is \emph{$\PT$-uniform} if the mapping $n \mapsto
  C_n$ is polynomial-time computable in $n$.
\end{definition}

>>>>

\subsection{Boolean Functions, Bases, and Circuits}
A (finite) \emph{Boolean function} is a function of the form $f: \{0,1\}^X
\rightarrow \{0,1\}$, for some finite index set $X$. We call a Boolean function
$f: \{0,1\}^X \rightarrow \{0,1\}$ \emph{symmetric} if for all $\sigma \in
\sym_X$ and $\vec{a} \in \{0,1\}^X$, $f (\vec{a} \sigma) = f(\vec{a})$. It
follows that a Boolean function is symmetric if, and only if, the output of the
function is entirely determined by the number of ones in the input. We say a
Boolean function $f : \{0,1\}^{*} \rightarrow \{0,1\}$ is \emph{symmetric} if
its restriction to inputs of size $n$ is symmetric for all $n \in \nats$.

It the literature a basis is usually taken to be a finite sequence of
(symmetric) functions from $\{0,1\}^*$ to $\{0,1\}$. In this paper we depart
slightly from this definition, and instead define a \emph{Boolean basis} (or
\emph{basis}) to be a (potentially infinite) set of finite Boolean functions. We
always denote a basis by $\BB$. We note that any function $f:\{0,1\}^* \ra
{0,1}$ is uniquely defined by a set of functions $\{f[n] : n \in \nats\}$, where
for all $n \in \nats$, $f[n]:\{0,1\}^n \ra \{0,1\}$ is the restriction of $f$ to
input strings of length $n$. It follows that any finite sequence of such
functions defines a basis given by taking the union of the associated sets of
finite functions. We denote the basis \emph{generated} by a sequence of
functions $f_1, \ldots, f_k : \{0,1\}^* \ra \{0,1\}$ by $\langle f_1, \ldots,
f_k \rangle := \bigcup_{i \in [k]} \{f_i[n] : n \in \nats\}$.

Given some of the generalisations used in this paper, we find it more convenient
to work over infinite sets of finite functions rather than the (equivalent)
unbounded fan-in functions. As such, we associate the usual logical operators
with infinite families of functions rather than a single unbounded fan-in
function. For each $n \in \nats$ we let $\AND[n], \OR[n], \NAND[n], \MAJ[n] :
\{0,1\}^n \rightarrow \{0,1\}$ be the usual logical operators, and with
$\MAJ[n]$ being the majority function for input strings of length $n$ (the
majority function outputs a one if, and only if, more than half of bits in the
inputs string are ones). We let $\AND := \{\AND[n] : n \in \nats\}$, and define
$\OR$, $\NAND$ and $\MAJ$ similarly. We write $\BS := \AND \cup \OR \cup \NAND
\cup \MAJ$ to denote the \emph{standard basis} and $\BM := \BS \cup \MAJ$ to
denote the \emph{majority basis}.

@@

\begin{definition}
  Let $(C_n)_{n \in \nats}$ be a family of Boolean circuits. We say that
  $(C_n)_{n \in \nats}$ is \emph{$\PT$-uniform} if the mapping $n \mapsto C_n$
  is polynomial-time computable in $n$.
\end{definition}

### 2018-03-10 sections/conclusion.tex
A complete draft minus the conclusion. Large changes made to a number of sections.
I have been trying to improve the presentation of the translation of circuits into formulas. To this end, I have changed the notation we use to talk about the usual logical operators (e.g. AND, OR, NOT, etc). I have also added tried to simplify the writing, and better explain the approach taken when translating circuits into formulas. I am in the process of writing the conclusion, but there is so much to think about.

\documentclass[../paper.tex]{subfiles}
\begin{document}
\end{document}

>>>>

\documentclass[../paper.tex]{subfiles}
\begin{document}

The main result of this paper establishes an equivalence between the expressive
power of $\FPR$ and the computational power of $\PT$-uniform transparent
symmetric rank-circuits, generalising a similar result by Anderson and
Dawar~\cite{AndersonD17}. Our motivation for proving this characterisation  The hope is that this characterisation will
contribute to our understanding of one of the two candidate logics for capturing
$\PT$.

% In the quest for a logic that captures polynomial time

% At the time of writing there are two candidate logics which Recent work on the
% problem of capturing $\PT$ has produced two `candidate' logics for capturing
% polynomial-time. In this paper we extend Anderson and Dawar's symmetric
% circuit characterisation of $\FPC$, showing that $P$-uniform families of
% transparent symmetric circuits with rank gates have the same expressive power
% as $\FPR$. Since $\FPR$ remains one of the two c the candidate logics for
% capturing $\PT$, along with choiceless polynomial time,
\end{document}

### 2018-03-10 sections/symmetric-circuits.tex
I have been trying to improve the presentation of the translation of circuits into formulas. To this end, I have changed the notation we use to talk about the usual logical operators (e.g. AND, OR, NOT, etc). I have also added tried to simplify the writing, and better explain the approach taken when translating circuits into formulas. I am in the process of writing the conclusion, but there is so much to think about.

% Let $\MB$ be the set of all structured functions that are symmetric or
% matrix-invariant.
We now discuss a few Boolean basis of interest in this paper and state the main
result. In most cases (e.g.\ \cite{}) a Boolean basis is taken to be a finite
sequence of symmetric Boolean functions. A binary encoding of a circuit defined
over such a basis is given by assigning to each element of the basis a unique
symbol, and then encoding the circuit in the obvious manner. Since there are
only a finite number of basis elements, it follows that if each basis element
can be computed in polynomial time then the mapping that takes a symbol $s$ and
a string $\vec{x}$ to the value of the function denoted by $s$ for input
$\vec{x}$ can be decided in polynomial-time. We call this function the
\emph{gate evaluation function}, and note that its polynomial-time decidability
is required for proving the polynomial-time evaluation of the circuit. However,
in order to allow for the inclusion of structured (possibly non-symmetric)
functions in the basis, we have instead defined a basis to be a (possibly
infinite) set of finite Boolean functions. In this case the gate evaluation
function may not even be decidable, let alone polynomial-time decidable.

\begin{definition}
  Let $\BB$ be a basis. Let $S$ be a set of symbols and let $F : \BB \rightarrow
  S$ be a bijection. We say $\BB$ is \emph{effective} if the function $E$ that
  maps a symbol $s$ and an input $x \in \dom(F(s))$ to the value
  $(F^{-1}(s))(x)$ is decidable. We say $\BB$ is \emph{$\PT$-effective} if $E$
  is polynomial-time decidable.
\end{definition}

@@

Let $n \in \nats$, we have that $\AND[n] : \{0,1\}^{[n]} \rightarrow \{0,1\}$ is
the restriction (as a structured function) of the $\AND$ function to inputs of
size $n$. Similarly, we have that $\OR[n]$, $\NAND[n]$ and $\MAJ[n]$ are the
restrictions of these respective functions to strings of length $n$. We let
$\NOT: \{0,1\} \ra \{0,1\}$ be the not function, and we often identify $\NOT$
with $\NAND[1]$. In each case we use the name of the function as the symbol
denoting that function.

As such, the standard basis $\BS$ consists of the functions $\NAND[n]$,
$\AND[n]$, and $\OR[n]$ for every $n \in \nats$, and the majority basis is $\BM
:= \BS \cup \{\MAJ[n] : n \in \nats\}$. We often use $\NAND[1]$ and $\NOT$
interchangeably. We say a basis $\BB$ is \emph{symmetric} if every function in
$\BB$ is a symmetric Boolean function.

Let $a, b, r, p \in \nats$, with $p$ prime. Let $\RANK^r_p[a,b] : \{0,1\}^{[a]
  \times [b]} \rightarrow \{0,1\}$ be a matrix-invariant structured function
with universe $[a] \uplus [b]$, such that $\RANK^r_p[a,b](M) = 1$ if, and only
if, the matrix $M \in \{0,1\}^{[a] \times [b]}$ has rank at most $r$ over $\ff_p$
when the entries of $M$ are interpreted as elements of $\mathbb{F}_p$. We
identify the function $\RANK^{r}_p[a,b]$ with the symbol $\RANK^{r}_p[a,b]$.

Let $\RANK = \{\RANK^{r}_p[a,b] : a, b, r, p \in \nats, \text{ $p$ prime}\}$ and
let the \emph{rank basis} be $\RB := \BM \cup \RANK$. We say a basis $\BB$ is
\emph{matrix-symmetric} if every function in $\BB$ is symmetric or
matrix-symmetric. It is easy to see that the standard basis, the majority basis,
and the rank basis are all $\PT$-effective.

@@

Let $C = \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a $(\mathbb{B},
\rho)$-circuit of order $n$ and let $g \in G$. We say a gate $g \in G$ is a
\emph{symmetric gate} if $\Sigma(g)$ is a symmetric function, and otherwise we
say $g$ is a \emph{non-symmetric gate} We say a gate $g \in G$ is a
\emph{matrix-symmetric gate} if $\Sigma(g)$ is a matrix-symmetric structured
function. We say that $g$ is an \emph{$\AND$-gate} if $\Sigma(g) := \AND[a]$ for
some $a \in \nats$. Similarly we define an \emph{$\OR$-gates},
\emph{$\NAND$-gate} \emph{$\NOT$-gates}, \emph{$\MAJ$-gates}, and
\emph{$\RANK$-gates}.

@@

\begin{thm}[Main Theorem]
  A graph property is decidable by a $P$-uniform family of transparent symmetric
  rank-circuits if, and only if, it is definable by an $\FPR$ sentence.
\end{thm}

@@

\subsection{Limitations of Symmetric Bases}
In this section we show that any family of symmetric circuits over an arbitrary
basis of symmetric functions can be transformed in polynomial time into a family
of symmetric circuits that decide the same language but are defined over the
basis $\BM$. In other words, it makes no difference to the expressive power of
such circuits whether we include \emph{all} symmetric functions in the basis or
just $\BM$. It follows that in order to construct $P$-uniform families of
symmetric circuits that define queires not in $\FPC$, we need to consider bases
with non-symmetric functions.

>>>>

% Let $\MB$ be the set of all structured functions that are symmetric or
% matrix-invariant.

Before we state the main result of this paper we need to discuss the Boolean
bases of interest in this paper and introduce some useful terminology. In most
other contexts a basis is taken to be a finite sequence of symmetric Boolean
functions, and a binary encoding of a circuit may is given by assigning to each
element of the basis a unique symbol, and then encoding the circuit as a
labelled DAG. Moreover, since there are only a finite number of basis elements,
it follows that if each function in the basis can be computed in polynomial time
then the mapping that takes a symbol $s$ and a string $\vec{x}$ to the value of
the function denoted by $s$ for input $\vec{x}$ can also be decided in
polynomial-time. We call this function the \emph{gate evaluation function}. It
is easy to see that the gate evaluation function is polynomial-time computable
if, and only if, the a given circuit can be evaluated in polynomial-time. It is
here our that our definition of a basis introduces a complication. In order to
allow for the inclusion of structured (possibly non-symmetric) functions in the
basis, we have instead defined a basis to be a (possibly infinite) set of finite
Boolean functions. In this case the gate evaluation function may not even be
decidable, let alone polynomial-time decidable.

\begin{definition}
  Let $\BB$ be a basis. Let $S$ be a set of symbols and let $F : \BB \rightarrow
  S$. We say $\BB$ is \emph{effective} if the function $E$ that maps a symbol
  $s$ and an input $x \in \dom(F(s))$ to the value $(F^{-1}(s))(x)$ is
  decidable. We say $\BB$ is \emph{$\PT$-effective} if $E$ is polynomial-time
  decidable.
\end{definition}

@@

We recall from Section~\ref{sec:background} that the standard basis $\BS$
consists of the functions $\NAND[n]$, $\AND[n]$, and $\OR[n]$ for every $n \in
\nats$, and that the majority basis $\BM := \BS \cup \{\MAJ[n] : n \in \nats\}$.
When encoding a circuit over either of these bases we abuse notation and use the
name of the function as the symbol denoting that function (e.g.\ the function
$\AND[n]$ is associated with the symbol $\AND[n]$).

Let $a, b, r, p \in \nats$, with $p$ prime. Let $\RANK^r_p[a,b] : \{0,1\}^{[a]
  \times [b]} \rightarrow \{0,1\}$ be a matrix-invariant structured function
with universe $[a] \uplus [b]$, such that $\RANK^r_p[a,b](M) = 1$ if, and only
if, the matrix $M \in \{0,1\}^{[a] \times [b]}$ has rank at most $r$ over
$\ff_p$ when the entries of $M$ are interpreted as elements of $\mathbb{F}_p$.
We associate the function $\RANK^{r}_p[a,b]$ with the symbol $\RANK^{r}_p[a,b]$.
Let $\RANK = \{\RANK^{r}_p[a,b] : a, b, r, p \in \nats, \text{ $p$ prime}\}$ and
let the \emph{rank basis} be $\RB := \BM \cup \RANK$. It is easy to see, given
the encoding of the basis, that the standard, majority and rank basis are all
$\PT$-effective.

We say a basis $\BB$ is \emph{symmetric} if every function in $\BB$ is symmetric
and we say $\BB$ is \emph{matrix-symmetric} if every function in $\BB$ is
symmetric or matrix-symmetric. Let $T_{\rank} := \{\AND, \NAND, \OR, \MAJ,
\RANK\}$. We say that $f \in \RB$ has \emph{type} $t \in T_{\rank}$ if $F \in
t$.

@@

Let $C = \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a $(\mathbb{B},
\rho)$-circuit of order $n$ and let $g \in G$. We say a gate $g \in G$ is a
\emph{symmetric gate} if $\Sigma(g)$ is a symmetric function, and otherwise we
say $g$ is a \emph{non-symmetric gate} We say a gate $g \in G$ is a
\emph{matrix-symmetric gate} if $\Sigma(g)$ is a matrix-symmetric structured
function. We say that $g$ has \emph{type} $t \in T_{\rank} \uplus \rho \uplus$
if $g$ is an input gate and $\Sigma (g) = t$ or $g$ is an internal gate and
$\Sigma (g)$ has type $t$. If $g$ has type $t$ we say $g$ is a $t$-gate (e.g.\
if $g$ has type $\AND$ we call it an $\AND$-gate). It is easy to see that the
function that maps a circuit and a gate to (a symbol denoting) the type of the
gate is polynomial-time computable.

@@

\begin{thm}[Main Theorem]
  A graph property is decidable by a $\P$-uniform family of transparent
  symmetric rank-circuits if, and only if, it is definable by an $\FPR$
  sentence.
\end{thm}

@@

\subsection{Limitations of Symmetric Bases}
In this section we show that any family of symmetric circuits over an arbitrary
basis of symmetric functions can be transformed in polynomial time into a family
of symmetric circuits that decide the same language but are defined over the
basis $\BM$. In other words, it makes no difference to the expressive power of
such circuits whether we include \emph{all} symmetric functions in the basis or
just $\BM$. It follows that in order to construct $\P$-uniform families of
symmetric circuits that define queires not in $\FPC$, we need to consider bases
with non-symmetric functions.
